Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field

### 2024-05-07 paper/ms.tex
update ms

{\it Age Accuracy and Quality Flagging}: The fact that only $\approx$90\% of the
rotation-based and lithium-based ages agree suggests that in cases of
disagreement, either one or both could be inaccurate.

{\it Old Ages}: $>$4\,Gyr problematic for gyro models, as calibrated.  Adding
$\log g$, and perhaps metallicity, could be two useful additional axes
along which to explore empirical gyro/isochrone fitting.

>>>>

Most ages reported in this study come from a single source of information: rotation.
Factors other than stellar age and mass can influence rotation rates of main sequence stars.
Binarity is the best example of this: even intermediate-separation ($\sim$10\,AU) binaries are biased
toward rotating anomalously
rapidly
\citep{Meibom_2007,2016AJ....152..115S,2019ApJ...871..174S,Gillen_2020,Bouma_2021}.
Metallicity may also be a second-order effect
\citep{2020MNRAS.499.3481A,2024arXiv240500779S}.
In this work, we were interested in not only the overall
age distribution of the Kepler field, but also in the reliability of ages for
individual young planet hosting stars.
The quality flagging scheme in Table~\ref{tab:planets} provides one method in support of
this end.
The other method is to consider the region in which we do have
overlapping sensitivity between lithium and rotation -- 
G and K dwarfs with $t$$\lesssim$1\,Gyr.  
Our results for such stars include
\ltonegyrhighqconfirmedtwosided\ planets orbiting ``high-quality''
stars for which the two age indicators are both constraining and in agreement
(with $Q_{\rm gyro}$ bits zero through nine not raised, and no planet quality flags raised).
Conversely, we found only one system, Kepler-786, with a clearly
constraining lithium age (\kepseveneightsix\,Myr) and no rotation age.

Other age indicators may help in verifying the ages of
the $\lesssim$3\,Gyr stars that are the focus of this work.
Specifically, chromospheric emission in the
X-ray, \ion{Ca}{2} HK line, and the UV can serve as an age tracer
\citep{2008ApJ...687.1264M,2014MNRAS.441.2361V,2024ApJ...960...62E}.
However, these age indicators are all in a sense ``rotation-powered''.  The dynamo
converts kinetic energy into magnetic energy, which is emitted through these chromospheric
pathways.
We did not attempt to incorporate these indicators into this study
due to concerns regarding sensitivity, homogeneity, and
the question of whether they in fact provide age
information that is truly independent from rotation.

A broader question, beyond the scope of this work, concerns the ages of
the older $\gtrsim$3\,Gyr stars that we do not treat.
For age-dating models based only on rotation rates, or else
on a combination of rotation and spin-modulation amplitude
\citep{2023ApJ...952..131M}, an important challenge is that the detectability
of the rotation signals for 3-10\,Gyr stars, especially Sun-like stars,
is at the limits of the current data \citep{2022ApJ...937...94M}.
In the ``old star'' regime, methods that use both the evolution of
stellar luminosity and stellar rotation \citep{Angus_2019} seem the most
capable of providing useful age constraints for single stars.
Incorporating kinematics \citep{2021AJ....161..189L} seems likely to remain
most useful for stellar ensembles, due to the stochastically of kinematic heating.
